---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

## EDA


```{r}
library(tidyverse)
library(haven)
library(colorspace)
library(ggridges)
library(ggthemes)
library(gtsummary)
library(tidyr)
library(broom)
library(ggplot2)
```

```{r}
# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, fig.retina = 2, dpi = 150, out.width = "60%"
)
# dplyr print min and max
options(dplyr.print_max = 10, dplyr.print_min = 10)
```

## DATA WRANGLING

```{r}
# Load assignment data that has characteristics of each tester. Might have multiple rows for a single tester
tester.data <- read_sas("../data/HDS_raw_data/assignment.sas7bdat", col_select = c("CONTROL", "TESTERID", "ARACE1", "TSEX"))
# Filter tester data to only African-Americans (AAs)
aa.tester.data <- tester.data %>% filter(ARACE1 == 2)

# Load skin color data. I'll use skin color to be consistent with the research paper that inspired the project
skin.color.data <- read_sas("../data/HDS_intermediate_data/skintone.sas7bdat")
# Inner join skin color data to tester data. Inner join because I don't want anyone without skin color data
aa.tester.data <- inner_join(aa.tester.data, skin.color.data, by = c("TESTERID" = "TesterID"))
```

```{r}
# Assess discrimination based on the following six outcomes that were also used in the paper. Two categories: monetary and non-monetary
# Some outcomes had flags that already made the comparison but others have to be coded here
outcome <- c(
  "CONTROL",
  "advrent_inc_min", # Rent for Advertised Unit Higher for Minority Tester  
  "advdepst_inc_min", # Deposit for Advertised Unit Higher for Minority Tester
  "rusptot_inc_wht", # Better Incentives for White Tester
  "totaval_inc_wht", # More Units Available to White Tester
  "min_rcredchk", # Agent Said Credit Check Required, Minority Tester
  "wht_rcredchk", # Agent Said Credit Check Required, White Tester
  "min_rcrimchk", # Agent Said Criminal Check Required, Minority Tester
  "wht_rcrimchk", # Agent Said Criminal Check Required, White Tester
  "site"
)

# Load intermediate rental data.  Each record represents one completed test that passed quality review.
# Only load the relevant columns as this is a very large file
rental.int.data <- read_sas("../data/HDS_intermediate_data/rental_intermediate.sas7bdat", col_select = all_of(outcome))
```

```{r}
# Create 2 flags that are equal to 1 if only the minority tester was subjected to discrimination. 
rental.int.data <- rental.int.data %>%
  mutate(min_only_credit = case_when(min_rcredchk == 1 & wht_rcredchk == 0 ~ 1, TRUE ~ 0),
         min_only_criminal = case_when(min_rcrimchk == 1 & wht_rcrimchk == 0 ~ 1, TRUE ~ 0))
```

```{r}
# Min-max normalization function
min_max_normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

# Create normalized composite score of discrimination that is the sum of all the dependent variables in the data scaled to a value between 0 and 1
rental.int.data <- rental.int.data %>%
  mutate(disc_score = rowSums(across(-c("CONTROL", "min_rcredchk", "wht_rcredchk", "min_rcrimchk", "wht_rcrimchk", "site")), na.rm = TRUE),
         disc_score_norm = min_max_normalize(disc_score), 
         discrimination = ifelse(disc_score_norm > 0, 1, 0)
         ) 
```

```{r}
# Inner join tester data to rental data 
aa.tester.data <- inner_join(aa.tester.data, rental.int.data, by = "CONTROL")
```

```{r}
# Create new fields for reconfigure sex and skin tone variable

# Remove rows where skin_color is 1 = Light because there are only 6 testers
aa.tester.data <- aa.tester.data %>%
  filter(Skintone != 1)

aa.tester.data <- aa.tester.data %>%
  mutate(sex = case_when(TSEX == 0 ~ "Female", TSEX == 1 ~ "Male"),
         skin_color = factor(Skintone, levels = c(2, 3, 4), labels = c("Somewhat Light", "Somewhat Dark", "Dark")))

```

**How many testers are there in my sample? By skin color and by
gender?** Not very balanced by sex AND skin color. More somewhat dark
and dark. More somewhat dark women and more dark men.

```{r}
# aa.tester.data %>% select(sex, skin_color) %>% tbl_summary(by = sex, percent = "row") %>% add_overall() %>% add_n()
aa.tester.data %>%  
tbl_cross(
    row = skin_color,
    col = sex,
    percent = "cell"
  ) 

```

**Summary table of score values**

```{r}
aa.tester.data %>%
  group_by(sex, skin_color) %>%
  summarise(mean_disc_score = mean(disc_score_norm, na.rm = TRUE),
            median_disc_score = median(disc_score_norm, na.rm = TRUE),
            min_score = min(disc_score_norm, na.rm = TRUE),
            max_score = max(disc_score_norm, na.rm = TRUE),
            .groups = "drop")
```

## PLOTS

```{r}
custom_theme <- theme_wsj() +
  theme(
    # Plot titles
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 10, margin = margin(b = 20)),
    
    axis.title.x = element_text(size = 10, margin = margin(t = 20)),
    axis.text.x = element_text(size = 8),
    
    axis.text.y = element_text(size = 8),
    
    # axis.line = element_line(arrow = arrow(type='closed', length = unit(5,'pt'))), 
    
    legend.position = "none"
  )

theme_set(custom_theme)
```

### SKIN COLOR AND DISCRIMINATION CHART

What question do I want this chart to answer? - Does the severity and
incidence of discrimination differ by skin tone

```{r}

p <- ggplot(aa.tester.data, aes(x = disc_score_norm, y = skin_color, fill = skin_color)) +
     geom_density_ridges(alpha = 0.8, color = "orange3") +
     scale_fill_manual(values = c("#FDE39C", "#f18d3f", "#7c4236")) + 
     labs(
         x = "Discrimination Score", 
         y = "Tester Skin Color",
         title = "Colorism and Rental Discrimination",
         subtitle = "Does a tester's skin color impact the discrimination they face?",
         caption = "Source: HUD Housing Discrimination Against Racial And Ethnic Minorities 2012 Dataset"
         ) 

p + theme(plot.caption = element_text(size = 5, color = "gray", margin = margin(t = 15))) +
  scale_x_continuous(limits = c(-0.19, 0.75), labels = c("Very Low","Low", "Moderate", "High", "Extreme"))

# ggsave("../plots/colorism-x-discrimination.png")

```

### SEX AND DISCRIMINATION CHART

What question do I want this chart to answer? - Which sex deals with
more severe discrimination?

```{r}
# Sex discrimination dumbbell plot

# Create summary table on sex
sex_df <-
  aa.tester.data %>%
  group_by(sex) %>%
  summarise(avg_score = mean(disc_score_norm, na.rm = TRUE),
            count = sum(discrimination, na.rm = TRUE),
            total = n(),
            prop = count/total)

sex_df %>% 
  pivot_longer(sex) %>%
  ggplot(aes(x=avg_score, y=name)) +
  geom_line(aes(group=name), color="gray", linewidth=3.5) +
  geom_point(aes(color=value), size=7) +
  theme(legend.position = "bottom") +
  scale_color_manual(values=c("#682d50", "#c4cc84")) +
   labs(
         x = "Discrimination"
   )

```

### SEX,SKIN COLOR AND DISCRIMINATION CHART

```{r}
# Create summary table on sex
sex_color_df <-
  aa.tester.data %>%
  group_by(sex, skin_color) %>%
  summarise(avg_score = mean(disc_score_norm, na.rm = TRUE),
            count = sum(discrimination, na.rm = TRUE),
            total = n(),
            prop = count/total,
            .groups = "drop")

# To gap or not to gap?
female_scores <- sex_color_df$avg_score[sex_color_df$sex == "Female"]
male_scores <- sex_color_df$avg_score[sex_color_df$sex == "Male"]
gap <- round(abs(male_scores - female_scores), 2)

sex_color_df %>%
  pivot_longer(skin_color) %>%  
  ggplot(aes(x=avg_score, y=value)) +
  geom_line(aes(group=value), color="gray", linewidth=3.5) +
  geom_point(aes(color=sex), size=7) + 
  geom_text(aes(label=sex, color=sex), data=. %>% filter(value=="Dark"), nudge_y =.2, fontface="bold") +
  scale_color_manual(values=c("#682d50", "#c4cc84")) +
  labs( title = "How do gender and skin color 
        impact discrimination?",
        x = "Mean Discrimination Score (0-1)") 




```



Do a chi squared to see if statistically significant

honeycomb monetarry and non monetary- color by gender and skin

```{r}
monetary_outcomes <- c( "advrent_inc_min", 
                        "advdepst_inc_min", 
                        "rusptot_inc_wht"
                        )

nonmonetary_outcomes <- c("totaval_inc_wht",
                          "min_only_credit",
                          "min_only_criminal"
                          )
aa.tester.data <- aa.tester.data %>%
  mutate(
         mon_disc_score = rowSums(across(all_of(monetary_outcomes)), na.rm = TRUE),
         mon_score_norm = min_max_normalize(mon_disc_score),
         nm_disc_score = rowSums(across(all_of(nonmonetary_outcomes)), na.rm = TRUE),
         nm_score_norm = min_max_normalize(nm_disc_score)
         )


aa.tester.data %>%
  group_by(skin_color) %>%
  summarise(mean_mon_score = mean(mon_score_norm, na.rm = TRUE),
            mean_nm_score = mean(nm_score_norm, na.rm = TRUE),
            .groups = "drop")
```

```{r}
ggplot(aa.tester.data, aes(x = mon_disc_score, y = nm_disc_score, color = skin_color, alpha = 0.8)) + 
  geom_jitter(size = 1.5) +
  facet_grid(cols = vars(skin_color), rows = vars(sex)) +
  scale_color_manual(values = c("Somewhat Light" = "#b19f6d", "Somewhat Dark" = "#f18d3f", "Dark" = "#7c4236"))

```

```{r}
# aa.tester.data %>% count(site, sort = TRUE)

northern_cities <- c("BT", "BN", "CH", "CL", "DT", "MS", "NO", "NT", "NY", "OC", "PD", "RS", "ST", "LA", "SD", "SJ")
southern_cities <- c("AT", "AQ", "CO", "DL", "FW", "GB", "HS", "MI", "SA", "TP", "KC", "RM")

aa.tester.data <- aa.tester.data %>% mutate(north_south = ifelse(site %in% northern_cities, "Northern", "Southern"))

table(aa.tester.data$north_south)

```

```{r}
ggplot(aa.tester.data, aes(x = mon_disc_score, y = nm_disc_score, color = skin_color, alpha = 0.8)) + 
  geom_jitter(size = 1.5) +
  facet_grid(cols = vars(skin_color), rows = vars(north_south)) +
  # theme(legend.position = "bottom", axis.title.y = element_text(size = 10)) +
  scale_color_manual(values = c("Somewhat Light" = "#b19f6d", "Somewhat Dark" = "#f18d3f", "Dark" = "#7c4236"))

```

```{r}
ggplot(aa.tester.data, aes(x = mon_disc_score, y = nm_disc_score, color = site)) + 
  geom_jitter() +
  facet_grid(cols = vars(skin_color), rows = vars(sex)) +
  theme(legend.position = "bottom") 
```


```{r}
aa.tester.data %>%
table(aa.tester.data$sex, aa.tester.data$skin_color)
```
```{r}
chisq.test(table(aa.tester.data$sex, aa.tester.data$skin_color))
```
```{r}
# Save aa tester data to use in shiny app
saveRDS(aa.tester.data, file = "shiny/aa_tester_data.rds")
```

